Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-17 paper.tex
progress

\newcommand{\Set}[1]{\mathsf{Set_{#1}}}
\newcommand{\Seti}{\mathsf{Set}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\Nat}{\mathsf{Nat}}
\newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\up}{\uparrow}
\newcommand{\down}{\downarrow}
\newcommand{\Lift}{\mathsf{Lift}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\Acc}{\mathsf{Acc}}
\newcommand{\acc}{\mathsf{acc}}
\newcommand{\Lvl}{\mathsf{Lvl}}

@@

@@

In the rest of the paper we make use of categories equipped with possibly
multiple family structures, which serves as basis for specifying universe
hierarchies. However, it is not very useful to simply have multiple copies of
family structures together with their type formers. In that case, every
constructor and eliminator of every type former stays in the same family, and
there is no interaction between families, the most we can do is to mix them
together in typing contexts. In this subsection we describe several ways of
crossing between families.

@@

\paragraph{Two-level type theory}
Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$), and a
lifting between them. This corresponds to a basic version of \emph{two-level
  type theory} \cite{twolevel}. This theory can be interpreted as having ($\Ty_1$,
$\Tm_1$) as a metaprogramming layer which can generate object-level programs in
the ($\Ty_0$, $\Tm_0$) layer.

@@

\begin{mydefinition} A \emph{family inclusion} is a lifting which preserves all type
and term formers. This assumes that every type former which is contained in the source family,
is also contained in the target family.
\end{mydefinition}

@@

\begin{mydefinition}\label{def:strict_inclusion}
A \emph{strict family inclusion} between ($\Tm_0$, $\Ty_0$) and ($\Tm_1$,
$\Ty_1$) is a family inclusion for which the following equations hold:
\begin{alignat}{2}
  & (\Gamma \ext \Lift\,A) &&= (\Gamma \ext A)     \label{eq:cumcon}    \\
  & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
  & \up\!t &&= t                                    \label{eq:cumlift}
\end{alignat}

%% \begin{enumerate}
%% \item $(\Gamma \ext \Lift\,A) = (\Gamma \ext A)$
%% \item $\Tm_1\,\Gamma\,(\Lift\,A) = \Tm_0\,\Gamma\,A$
%% \item $\up\!t = t$
%% \end{enumerate}
\end{mydefinition}

@@

The definitional equality of $\true_0$ and $\true_1$ is important; without it
canonicity would fail, since $\true_0$, $\false_0$, $\true_1$ and $\false_1$
would be four definitionally distinct inhabitants of $\Bool_1$. See Luo
\cite{luo2012notes} on related issues with cumulativity. This means that it is
not sufficient to specify a strict inclusion just by equations \ref{eq:cumcon}
and \ref{eq:cumtm} in Definition \ref{def:strict_inclusion}, we need $\up$ together
with equation \ref{eq:cumlift} to identify term formers in different families. The
other direction $\down\!t = t$ is immediately derivable.

@@

\begin{mylemma} Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are propositionally equal; see
  \cite[Section 10.3]{hottbook}.
\end{mylemma}

@@

@@

\begin{mydefinition}
A \emph{family diagram} over $\Lvl$ maps each $i : \Lvl$ to a family structure
($\Ty_i$, $\Tm_i$), and each $p : i < j$ to a strict inclusion
($\Lift_{i}^{j}\,p$, $\up_{i}^{j}p$, $\down_{i}^{j}p$) between ($\Ty_i$,
$\Tm_i$) and ($\Ty_j$, $\Tm_j$). Moreover, the mapping is functorial, so
$\Lift_{i}^{k}\,(p\circ q)\,A = \Lift_{j}^{k}\,p\,(\Lift_{i}^{j}\,q\,A)$, and
similarly for $\up_{i}^{j}p$ and $\down_{i}^{j}p$.
\end{mydefinition}

@@

Our choice of level structures and diagrams is motivated by the following
reasons. First, we do not need identity morphisms in levels, because they would
be mapped to trivial liftings, which are not interesting in our setting. Second,
we do not need proof-relevant level morphisms, since any parallel pair of family
liftings gives rise to isomorphic types. Concretely, given $p : i < j$ and $q :
i < j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq
\Tm_i\,\Gamma\,A \simeq \Tm_j\,\Gamma\,(\Lift\,q\,A)$, and since $\Lift\,p\,A$
and $\Lift\,q\,A$ are in the same family, we can internally prove them
isomorphic using function types and identity types.

%% \begin{notation}
%% Given $p : i < j$ we may write $\Lift_{i}^{j}\,p : \Ty_i\,\Gamma \to
%% \Ty_j\,\Gamma$ for type lifting, and $\up_{i}^{j}p$ and $\down_{i}^{j}p$
%% accordingly for term lifting. Sometimes we omit $i$, $j$ and $p$ annotations,
%% when there is no ambiguity.
%% \end{notation}

%% \end{definition}

@@

Coquand, Russell, naturality, large elimination examples
Lifting of universes
No lifting naturality for (El,Code)!

@@

>>>>

\newcommand{\Set}[1]{\mathsf{Set_{#1}}}
\newcommand{\Seti}{\mathsf{Set}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\Nat}{\mathsf{Nat}}
\newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\up}{\uparrow}
\newcommand{\down}{\downarrow}
\newcommand{\Lift}{\mathsf{Lift}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\Acc}{\mathsf{Acc}}
\newcommand{\acc}{\mathsf{acc}}
\newcommand{\Lvl}{\mathsf{Lvl}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}

@@

In the following, whenever we introduce a type former, we always assume that it
is natural with respect to substitution, i.e.\ all type and term formers have a
corresponding substitution rule.

@@

In the rest of the paper we make use of categories equipped with possibly
multiple family structures, which serves as basis for specifying universe
hierarchies. However, it is not very useful to simply have multiple copies of
family structures together with their type formers. In that case, every
constructor and eliminator of every type former stays in the same family, and
there is no interaction between families, and the most we can do is to mix them
together in typing contexts. In this subsection we describe several ways of
crossing between families.

@@

\paragraph{Two-level type theory}
Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a
lifting between them. This corresponds to a basic version of \emph{two-level
  type theory} \cite{twolevel}. This theory can be interpreted as having ($\Ty_1$,
$\Tm_1$) as a metaprogramming layer which can generate object-level programs in
the ($\Ty_0$, $\Tm_0$) layer.

@@

\begin{mydefinition}\label{def:inclusion}
A \emph{family inclusion} is a lifting which preserves all type and term
formers. This assumes that every type former which is contained in the source
family, is also contained in the target family.
\end{mydefinition}

@@

\begin{mydefinition}\label{def:strict_inclusion}
A \emph{strict family inclusion} between ($\Tm_0$, $\Ty_0$) and ($\Tm_1$,
$\Ty_1$) is a family inclusion ($\Lift$, $\up$, $\down$) for which the following equations hold:
\begin{alignat}{2}
  & (\Gamma \ext \Lift\,A) &&= (\Gamma \ext A)     \label{eq:cumcon}    \\
  & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
  & \up\!t &&= t                                    \label{eq:cumlift}
\end{alignat}
\end{mydefinition}

@@

The definitional equality of $\true_0$ and $\true_1$ is important; without it
canonicity would fail, since $\true_0$, $\false_0$, $\true_1$ and $\false_1$
would be four definitionally distinct inhabitants of $\Bool_1$. See Luo
\cite{luo2012notes} for a discussion of related issues with cumulativity. It is
not sufficient to specify a strict inclusion just by equations \ref{eq:cumcon}
and \ref{eq:cumtm} in Definition \ref{def:strict_inclusion}, we need $\up$
together with equation \ref{eq:cumlift} to identify term formers in different
families. The other direction $\down\!t = t$ is immediately derivable.

@@

\begin{mylemma}
Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are
propositionally equal \cite[Lemma 10.3.4]{hottbook}.
\end{mylemma}

@@

\begin{mylemma}
Every well-founded relation is irreflexive \cite[Lemma 10.4.2]{hottbook}.
\end{mylemma}

@@

\begin{mydefinition}
A \emph{family diagram} over $\Lvl$ maps each $i : \Lvl$ to a family structure
($\Ty_i$, $\Tm_i$), and each $p : i < j$ to a family inclusion
($\Lift_{i}^{j}\,p$, $\up_{i}^{j}p$, $\down_{i}^{j}p$) between ($\Ty_i$,
$\Tm_i$) and ($\Ty_j$, $\Tm_j$). Moreover, the mapping is functorial, so
$\Lift_{i}^{k}\,(p\circ q)\,A = \Lift_{j}^{k}\,p\,(\Lift_{i}^{j}\,q\,A)$, and
similarly for $\up_{i}^{j}p$ and $\down_{i}^{j}p$. A \emph{strict family diagram}
is a family diagram where each inclusion is strict.
\end{mydefinition}

@@

Our choice of level structures and diagrams is motivated by the following
reasons. First, we do not need identity morphisms in levels, because they would
be mapped to trivial liftings, which are not interesting in our setting. Second,
we do not need proof-relevant level morphisms, since any parallel pair of family
liftings gives rise to isomorphic types. Concretely, given $p : i < j$ and $q :
i < j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq
\Tm_i\,\Gamma\,A \simeq \Tm_j\,\Gamma\,(\Lift\,q\,A)$, and since $\Lift\,p\,A$
and $\Lift\,q\,A$ are in the same family, we can internally prove them
isomorphic using function types and identity types. That said, every construction
in this paper would still work with inverse categories as level structures.

@@

At this point, we can talk about family diagrams, but no previously seen type
former depends on levels in an interesting way. For example, $\Bool_i$ has the
same canonical inhabitants as $\Bool_j$, for any $i$ and $j$. Universes
introduce dependency on levels, by serving as classifiers for smaller families internally
to larger families.

\todo{find reference for Coquand univ}

\begin{mydefinition} A family diagram supports \emph{universe formation} if it supports the following:
\begin{alignat*}{3}
  & \U             &&: (i\,j : \Lvl) \to i < j \to \Ty_j\,\Gamma\\
  & \mathsf{LiftU} &&: \Lift_{j}^{k}\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)
\end{alignat*}
\end{mydefinition}
We also need a way to pin down universes as classifiers. We consider two variants.
\begin{mydefinition}
A family diagram has \emph{Coquand universes} if it has universe formation and
additionally supports $\El : \Tm_j\,\Gamma\,(\U\,i\,j\,p) \to \Ty_i\,\Gamma$, and
its inverse $\Code : \Ty_i\,\Gamma \to \Tm_j\,\Gamma\,(\U\,i\,j\,p)$.
\end{mydefinition}
\begin{mydefinition}
A family diagram has \emph{Russell universes} if it has Coquand universes and
additionally satisfies $\Tm_j\,\Gamma\,(\U\,i\,j\,p) = \Ty_i\,\Gamma$ and
$\El\,t = t$.
\end{mydefinition}

The move from Coquand to Russell universes is fairly similar to the move from
inclusions to strict inclusions. The Russell variant makes it possible to
informally omit $\El$ and $\Code$. Likewise, the $\El\,t = t$ condition ensures
appropriate naturality. If we only assumed $\Tm_j\,\Gamma\,(\U\,i\,j\,p) =
\Ty_i\,\Gamma$ but not Coquand universes, we would not be able to prove that a
$t : \Tm_j\,\Gamma\,(\U\,i\,j\,p)$ substituted \emph{as a term} is the same
thing as $t$ substituted \emph{as a type}. Both would be written as $t[\sigma]$
in our notation, but they involve different $\blank[\blank]$ operations.

Unlike every other type or term former, there is no lifting computation rule for
$\El$ and $\Code$. Intuitively, the issue is that we would need to relate type
lifting and term lifting, but while term lifting has invertible action, type
lifting does not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and a
larger $\Ty$ is not isomorphic a smaller one, because it contains more
universes. So, for example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to
$\Ty_1\,\Gamma$ yields $\Bool_1$, but lifting $\Bool_0$ as a term yields
$\Bool_0$.

Assuming Coquand or Russell universes and $p : i < j$, we can recover
polymorphic functions, for example, we may have $\mi{id} : \Pi(A : \U\,i\,j\,p)
(\Lift\,p\,(\El\,A) \to \Lift\,p\,(\El\,A))$ for the polymorphic identity. Here,
we quantify over terms of $\U$, and since every type former stays on the same
level (including $\Pi$), we have to $\Lift$ the types in the codomain to the
level of the domain. We can also recover large elimination, for example as in
\[
(\lambda\,(b : \Bool_j).\,\mathsf{if}\,b\,\mathsf{then}\,\Code\,\top_i\,\mathsf{else}\,\Code\,\bot_i)
: \Tm_j\,\Gamma\,(\Bool_j \to \U\,i\,j\,p).
\]

@@

In this section we give a model for a type theory with generalized
universes. Let's make the notion of model concrete first.

\begin{mydefinition} Fix a $\Lvl$ structure. A model for a type theory with $\Lvl$-indexed
universes consists of the following:
\begin{enumerate}
  \item A base category ($\Con$, $\Sub$) with a terminal object $\emptycon$.
  \item A strict family diagram ($\Ty_i$, $\Tm_i$) over $\Lvl$, supporting Russell
        universes, and each family structure is closed under the same basic type formers.
\end{enumerate}
\end{mydefinition}

We abbreviate the theory as TTGU, for ``type theory with generalized
universes''. The choice of available basic type formers is up to personal taste,
and it will not significantly affect the following model construction.
